# Population {#population}

```{r}
#| label: setup

library(tidyverse)
library(tidycensus)
library(janitor)

```

The dashboards on this page include data on total population counts and components of population change. These trends help reveal how much, and for what reasons, population levels may be increasing or decreasing over time.

## Total Population

This dashboard shows annual population counts and percent changes from 2010 to 2022. Data for 2010 and 2020 are from those years' Decennial Census counts; all other years use estimates from the Census Bureau's Population Estimates Program (PEP).

```{r}
#| label: pop-data
#| eval: false

# Download total population estimates from PEP for each Virginia locality from 2010 to 2019

pep_2010s <- get_estimates(
  geography = "county",
  state = "VA",
  variables = "POP",
  year = 2019,
  time_series = TRUE
) |> 
  filter(!DATE %in% c(2, 3)) |> # Remove non-Decennial 2010 estimates
  mutate(year = # Translate date codes into years
    case_when(
      DATE == 1 ~ "2010",
      DATE == 4 ~ "2011",
      DATE == 5 ~ "2012",
      DATE == 6 ~ "2013",
      DATE == 7 ~ "2014",
      DATE == 8 ~ "2015",
      DATE == 9 ~ "2016",
      DATE == 10 ~ "2017",
      DATE == 11 ~ "2018",
      DATE == 12 ~ "2019")) |> 
   mutate(counttype = # Add descriptions to count types
      case_when(
        DATE == 1 ~ "Census population",
        TRUE ~ "Population estimate")) |>  
  select(GEOID, counttype,year, value) # Simplify data

# Download total population estimates from PEP for each Virginia locality from 2021 to 2022.

pep_2020s <- get_estimates(
  geography = "county",
  state = "VA",
  variables = "POPESTIMATE",
  year = 2022,
  time_series = TRUE
) |> 
  filter(year != 2020) |> # Remove non-Decennial 2020 count
  mutate(counttype = case_when( # Add descriptions to count types
    variable == "POPESTIMATE" ~ "Population estimate",
    TRUE ~ variable
  )) |> 
  select(GEOID,counttype, year, value) # Simplify data

# Download total population counts from 2020 Decennial Census.

census <- get_decennial(
  geography = "county",
  state = "VA",
  year = 2020,
  sumfile = "pl",
  variables = "P1_001N"
) |> 
  mutate(counttype = "Census population",
         year = 2020) |> 
  select(GEOID, counttype, year, value)

# Combine data frames into one and join to lookup table.

pop_data <- rbind(pep_2010s, pep_2020s, census) # Combine data frames

lookup <- read_csv("data/local_lookup.csv") |> # Read in lookup csv
  mutate(fips_full = as.character(fips_full)) |> # Convert numeric GEOID to character in order to complete join
  select(GEOID = fips_full, name_long, cbsa_title) # Simplify data

pop_data_join <- pop_data |> 
  left_join(lookup, by = 'GEOID')

# Write data to rds format in data folder and note the date below.

# Data up-to-date as of: 5-22-2023

write_rds(pop_data_join, "data/total_pop.rds")
write_rds(pop_data_join, "shiny/total_pop/total_pop.rds")

```

## Components of Population Change

This dashboard breaks down annual population changes by three different components:

- Natural increases (births minus deaths),
- Domestic migration (net moves from in/out of state), and
- International migration (net moves from in/out of the country).

```{r}
#| label: change-data
#| eval: false

change_2010s <- get_estimates(
  geography = "county",
  state = "VA",
  variables = c("NATURALINC", "DOMESTICMIG", "INTERNATIONALMIG"),
  year = 2019,
  time_series = TRUE
) |> 
  mutate(year = # Translate date codes into years
    case_when(
      PERIOD == 1 ~ "2010",
      PERIOD == 2 ~ "2011",
      PERIOD == 3 ~ "2012",
      PERIOD == 4 ~ "2013",
      PERIOD == 5 ~ "2014",
      PERIOD == 6 ~ "2015",
      PERIOD == 7 ~ "2016",
      PERIOD == 8 ~ "2017",
      PERIOD == 9 ~ "2018",
      PERIOD == 10 ~ "2019")) |> 
  select(GEOID, variable, year, value)
  

change_2020s <- get_estimates(
  geography = "county",
  state = "VA",
  variables = c("NATURALCHG", "DOMESTICMIG", "INTERNATIONALMIG"),
  year = 2022,
  time_series = TRUE
) |> 
  select(GEOID, variable, year, value)

change_data <- rbind(change_2010s, change_2020s) |> 
  mutate(component = # Rename components of change
    case_when(
      variable == "NATURALINC" ~ "Natural increase",
      variable == "NATURALCHG" ~ "Natural increase", 
      variable == "DOMESTICMIG" ~ "Domestic migration",
      variable == "INTERNATIONALMIG" ~ "International migration"))

lookup <- read_csv("data/local_lookup.csv") |> # Read in lookup csv
  mutate(fips_full = as.character(fips_full)) |> # Convert numeric GEOID to character in order to complete join
  select(GEOID = fips_full, name_long, cbsa_title) # Simplify data

change_data_join <- change_data |> 
  left_join(lookup, by = 'GEOID')

# Write data to rds format in data folder and note the date below:

# Data up-to-date as of: 5-22-23

write_rds(change_data_join, "data/pop_change.rds")

# Shiny Dashboard should show the changes over time utilizing a faceted bar chart for 2010 through 2022. Three different data viz should show data by state, CBSA, and locality. Filters are on the latter two, which allow you to switch to different geographies.

```



### Additional resources

**U.S. Census Bureau**

[*Virginia adds more than 600,000 people since 2010*](https://www.census.gov/library/stories/state-by-state/virginia-population-change-between-census-decade.html)

**UVA Weldon Cooper Center**

[*State population growth slows, with most new residents in Northern Virginia*](https://news.virginia.edu/content/state-population-growth-slows-most-new-residents-northern-virginia)

[*Comparing 2020 Virginia population projections to the Census count*](https://statchatva.org/2021/09/27/comparing-2020-virginia-population-projections-to-the-census-count/)

## Data sources

[Census Population Estimates Program](https://www.census.gov/programs-surveys/popest.html) (PEP)

Annual estimates of total population released by the U.S. Census Bureau. Available at locality level from 2010 to 2019. Estimates by age, race, and ethnicity also available, along with components of population change.

[2020 Decennial Census](https://www.census.gov/programs-surveys/decennial-census/about/rdo/summary-files.html)

Initial data from the 2020 Decennial Census is available in the PL 94-171 Redistricting summary file. As of February 2022, we use this data to append 2020 population counts onto the PEP time series data through 2019. Complete 2010 to 2020 Intercensal Estimates will be published in late 2022.
